% \section{Router Configuration}

%\begin{multicols}{2}

The names (determined at compile-time) and interface specifications for the
routing and crossbar functions, are the only crucial ``magical'' things one
needs to contend with in a proper {\em router} configuration.  The syntax and
semantics of the configuration file's contents are dealt with in the
following subsection. The details of the two functions introduced here are
specified after that, once the necessary background information has been
given.

{\em Router} behavior is controlled by a configuration file read at 
startup. It is a {\tt zmsh(1)} script that uses facilities provided 
built into the {\em router}. 

The configuration file looks like a Bourne Shell script at first glance.
There are minor syntax changes from standard {\tt sh}, but the aim is to be as
close to the Bourne Shell language as is practical. The contents of the
file are compiled into a parse tree, which can then be interpreted by the
{\em router}.  The configuration file is usually self-contained, although an easy
mechanism exists to make use of external UNIX programs when so desired.
Together with a very flexible database lookup mechanism, functions, and
address manipulation based on token-matching regular expressions, the
configuration file language is an extremely flexible substrate to
accomplish its purpose. When the language is inadequate, or if speed
becomes an issue, it is possible to call built in (C coded) functions. The
interface to these functions is mostly identical to what a standalone
program would expect (modulo symbol name clashes and return values), to
ease migration of external programs to inclusion in the {\em router} process.




\subsection{Configuration File Programming Language}



Whenever the {\em router} process starts, its first action is to read its
configuration file.  The configuration file is a text file which contains
statements interpreted immediately when the file is read.  Some statements
are functions, in which case the function is defined at that point in
reading the configuration file.  The purpose of the configuration file is
to provide a simple way to customize the behavior of the mailer, and this
is primarily achieved by defining the {\tt router} and {\tt crossbar} 
functions. For these to work properly, some initialization code and auxiliary
functions will usually be needed.

At first sight, a configuration file looks like a Bourne shell script.
The ideal is to duplicate the functionality, syntax, and to a large
degree the semantics, of a shell script.  Therefore, the configuration file
programming language is defined in terms of its deviation from standard
Bourne shell syntax and semantics.  The present differences are:

\begin{itemize}
\item
No {\tt repeat} statement.
\item
Functions are allowed, parameter lists are allowed. If not enough
arguments are present in a function call to exhaust the parameter
list, the so-far unbound parameter variables are bound to `' (the
empty string) as local variables. For example, this is the identity
address rewriting function:
     
\begin{alltt}
null (address) \{
  return $address  # surprise!
\}
\end{alltt}

\item
Multiple-value returns are allowed.
The {\tt return} statement can be used to return a non-``'' value
from a function.
The following are all legal {\tt return} statements:
\begin{alltt}\medskip
return
return $address
return $channel ${next_host} ${next_address}
\end{alltt}

\item
Variables are dynamically scoped, local variables are the ones
in a function's parameter list and those declared with the ``{\tt local}''
statement.
Only the first value of a multiple-value return may be assigned to a variable.
All values are strings, so no type information, checking, or declaration,
is  necessary.
\item
Quoting is a bit stilted.
All quotes (double-, single-, back-), must appear in matching pairs at
the beginning and at the end of a word.

{\bf\large CHECK!}
Single quotes are not stripped, double quotes cause the enclosed character
sequence to be collected into a quoted-string RFC822 token.

For example, the statement:
\begin{alltt}
 foo `bar "`baz`"`
\end{alltt}
is evaluated as
\begin{alltt}
 (apply 'foo (apply 'bar (baz)))
\end{alltt}

\item
Conditional substitution forms are supported:
\begin{alltt}
  $\{variable:=value\}
  $\{variable:-value\}
  $\{variable:+value\}
\end{alltt}

\item
Patterns (in case labels) are evaluated once,
the first time they are encountered.

\item
At the end of a case label, the sequentially next case labels of
the same case statement will be tried for successful pattern matching
(and the corresponding case label body executed).
The only exceptions (apart from encountering a return statement) are:
\begin{description}
\item[\tt again] \mbox{} \\
a function which retries the current case label for a match
\item[\tt break] \mbox{} \\
continues execution after the current case statement
\end{description}

\item
A regular expressions using variant of ``case'', with two
flavours:
\begin{description}
\item[ssift]
A ``String Sift'' where the input string is handled as is.
\item[tsift]
A ``Token Sift'' where the input string is spliced according to
RFC-822 tokenization rules.  Especially RFC-822 special characters
cause tokens to split.

With ``tsift'' the ``.'' (dot) becomes match any single {\bf token},
that is, input string ``foo.bar'' has three tokens: ``foo'', ``.'', and
``bar''.
\end{description}

Overall usage of these ``sifts'' is very much like that of ``case'',
including the need for matching termination tokens.
\begin{alltt}\medskip
  ssift "$invar" in
  pattern
        statements
        ;;
  tfiss
  tsift "$invar" in
  pattern
        statements
        ;;
  tfist
\end{alltt}

\item
Various standard Bourne shell functions do not exist built in.
\item
The general form of function calls in the system is:
\begin{alltt}
  \$(funcname arguments)
\end{alltt}
It returns a scalar or list object, and the result can be stored
into variables at will.

\item
Relations, and other database lookups are constructed as function calls
where the relation name is the function name.
More about this later.
\end{itemize}

There are currently only three entry points (i.e. magic names known to the
{\em router} code) in the configuration scripts, namely the {\tt process}, {\tt router} and 
{\tt crossbar} functions.

The {\tt process} function is called with a file name as argument. 
The file is typically located in the {\tt \$POSTOFFICE/router/} 
directory. {\tt Process} is a protocol switch function which uses the 
form of the file name to determine how to process different types of 
messages.

The {\tt router} function is called with an address as argument,
and returns a quad of (channel, host, user, attribs) as three
separate values,  corresponding to the channel the message should
be sent out on (or, the router function can also be called to check
on who sent a message), the host or node name for that channel
(semantics depend per what channel is in effect), and the address
the receiving agent should transmit to.
The fourth parameter is ``attribute'' storage variable name from
which a ``privilege'' value-pair is picked for recipient address
security control functions.

The {\tt crossbar} function is in charge of rewriting envelope addresses,
selecting message header address munging type (a function to be called with
each message header address), and possibly doing per-message logging or
enforcing restrictions deemed necessary.
It takes a sender-quad and a receiver-quad as arguments (eight parameters
altogether).
It returns the new values for each element of the two quads, and in addition
a function name corresponding to the function to be used to rewrite header
addresses for the specific destination.
If the destination is to be ignored, returning a null function name will
accomplish this.

The {\em router} has several built in (C coded) functions.
Their calling sequence and interface specification is exactly the same
as for the functions defined in the configuration file.
Some of these functions have special semantics, and they fall into three
classes, as follows:

\begin{itemize}
\item
Functions that are critical to the proper functioning of the configuration
file interpreter:

\begin{description}
\item[{\tt return}] \mbox{}

returns its argument(s) as the value of a function call

\item[{\tt again}] \mbox{}

repeats the current case, and *sift label

\item[{\tt break}] \mbox{}

exits case, and *sift statements

\end{description}


\item
Functions that are necessary to complete the capabilities of the
interpreter:

\begin{description}
\item[{\tt relation}] \mbox{}

defines a database to the database lookup mechanism

\item[{\tt sh}] \mbox{}

an internal function which runs its arguments as {\tt /bin/sh} would

\end{description}


\item
Non-critical but recommended functions:

\begin{description}
\item[{\tt getzenv}] \mbox{}

retrieves global ZMailer configuration values

\item[{\tt echo}] \mbox{}

emulates {\tt /bin/echo}

\item[{\tt exit}] \mbox{}

aborts the {\em router} with the specified status code

\item[{\tt hostname}] \mbox{}

internal function to get and set the system name

\item[{\tt trace}] \mbox{}

turns on selected debugging output

\item[{\tt untrace}] \mbox{}

turns off selected debugging output

\item[{\tt [}] \mbox{}

emulates a subset of {\tt /bin/test}
(a.k.a. \verb!/bin/[!) functionality.

\end{description}
\end{itemize}


The {\tt relation} function is described in 
{\em Databases}, section \vref{adm:Databases}.
Functions {\tt trace} and {\tt untrace} are described
in connection with debugging. 

See {\em Logging and Statistics},  section \vref{Logging_and_Statistics}.
({\bf This will probably change to Reference, Router, Debugging})

The {\tt hostname} function requires some further explanation.
It is intended to emulate the BSD UNIX {\tt /bin/hostname}
functionality, except that setting the hostname will only set
the {\em router's} idea of the hostname, not the system's.
Doing so will enable generation of `Message-Id' and `Received'
``trace'' headers on all messages processed by the {\em router}.
It is done this way since the {\em router} needs to know the official
domain name of the local host in order to properly generate these headers,
and this method is cleaner than reserving a magic variable for the purpose.
The {\em router} cannot assume the hostname reported by the system is
a properly qualified domain name, so the configuration file may generate
it using whichever method it chooses.

If the hostname indeed is a fully qualified domain name, then:
\begin{alltt}\medskip\hrule\medskip
  hostname "hostname"
\medskip\hrule\medskip\end{alltt}

will enable generation of trace headers.

Finally, note that a symbol can have both a function-value and a
string-value.
The string value is of course accessed using the \$ prefix
convention of the Bourne shell language.

{\em However careless usage of {\tt\$var} is DANGEROUS, as you
may trip on following:
\begin{alltt}\medskip\hrule\medskip
  SS="1 -o bar"
  if [ $SS = 0 ]; then
    ...
  fi
\medskip\hrule\end{alltt}\medskip
In this kind of places you must tread carefully, and use double-quotes
to save your neck:
\begin{alltt}\medskip\hrule\medskip
  SS="1 -o bar"
  if [ "$SS" = 0 ]; then
    ...
  fi
\medskip\hrule\end{alltt}\medskip
}

To test the configuration or routing data, proceed as
shown in figure \vref{fig:adm:testing}.

\begin{figure*}
\begin{alltt}\hrule\medskip
sh$ $MAILBIN/router -i        {\rm(select interactive mode)}
z$ rtrace                     {\rm(turn tracing on)}
z$ router user@broken.address {\rm(the address that gave you trouble)}
z$ router another@address     {\rm(and so on)}
\medskip\hrule\end{alltt}\medskip
\caption{Example of running tests on {\em router}}
\label{fig:adm:testing}
\end{figure*}

Old salts can use {\tt /usr/lib/sendmail -bt} instead of {\tt router -i}.
Once satisfied that routing works, command:
\begin{alltt}\medskip\hrule\medskip
  zmailer router
\medskip\hrule\end{alltt}\medskip
will restart the {\em router}.

You can also run the {\em router} directly on a message.  Copy your message
to someplace other than the postoffice ({\tt /tmp} is usually good), in a
numeric file name.  If the file name is {\tt 123}, you run
\begin{alltt}\medskip\hrule\medskip
  $MAILBIN/router 123
\medskip\hrule\medskip\end{alltt}
this will create the file ``{\tt .123}'' containing the control information
produced by the {\em router}.


\subsection{Databases}
\label{adm:Databases}

Many of the decisions and actions taken by configuration file code depend
on the specifics of the environment the MTA finds itself in.  So, not just
the facts that the local host is attached to (say) the UUCP network and a
Local Area Network are important, but it is also essential to know the specific
hosts that are reachable by this method.  Hardcoding large amounts of such
information into the configuration file is not practical.  It is also
undesirable to change what is really a program (the configuration file),
when the information (the data) changes.

The desirable solution to this data abstraction problem is to provide a way
for the configuration file programmer to manage such information externally
to ZMailer, and access it from within the {\em router}.  The logical way to do
this is to have an interface to externally maintained databases.  These
databases need not be terribly complicated; after all the simplest kind of
information needed is that a string is a member of some collection.  This
could simply correspond to finding that string as a word in a list of
words.

However, there are many ways to organize databases, and the necessary
interfaces cannot be known in advance.  The {\em router} therefore implements a
framework that allows flexible interfacing to databases, and easy extension
to cover new types of databases.

To use a database, two things are needed: the name of the database, and a
way of retrieving the data associated with a particular key from that
database.  In addition to this knowledge, the needs of an MTA do include
some special processing pertinent to its activities and the kind of keys to
be looked up.

Specifically, the result of the data lookup can take different forms: one
may be interested only in the existence of a datum, not its value, or one
may be looking up paths in a pathalias database and need to substitute the
proper thing in place of `{\tt \%s}' in the string returned from the database
lookup.  It should be possible to specify that this kind of postprocessing
should be carried out in association with a specific data access.
Similarly, there may be a need for search routines that depend on the
semantics of keys or the retrieved data.  These possibilities have all been
taken into consideration in the definition of a relation.  A relation maps
a key to a value obtained by applying the appropriate lookup and search
routines, and perhaps a postprocessing step, applied to a specified
database that has a specified access method.

The various attributes that define a relation are largely independent.
There will of course be dependencies due to the contents or other semantics
of a database.  In addition to the features mentioned, each relation may
optionally have associated with it a subtype, which is a string value used
to tell the lookup routine which table of several in a database
one is interested in.

There are no predefined relations in the {\em router}.
They must all be specified in the configuration file before first use.
This is done by calling the special function {\tt relation} with
various options, as indicated by the usage strings printed by
the {\tt relation} function when called the wrong way.
See figure \vref{fig:relation:params}.

\begin{figure*}
\begin{alltt}\medskip\hrule\medskip
Usage: relation -t dbtype[,subtype] [-f file -e# -s# -bilmnpu -d driver] name
  dbtypes: incore,header,unordered,ordered,hostsfile,bind,selfmatch,ndbm,gdbm,btree,bhash
\medskip\hrule\end{alltt}\medskip
\caption{\label{fig:relation:params}``Usage:'' report from {\tt relation}} 
\end{figure*}

The `{\tt t}' option specifies one of several predefined database types,
each with their specific lookup routine.
It determines a template for the set of attributes associated with
a particular relation.
The predefined database types are:


\begin{description}
\item[{\tt bhash}] \mbox{}

the database is in BSD DB HASH format.

\item[{\tt bind}] \mbox{}

the database is the BIND nameserver, accessed through the standard resolver routines.

\item[{\tt btree}] \mbox{}

the database is in BSD DB BTREE format.

\item[{\tt dbm}] \mbox{}

the database is in DBM format.
Note that the original dbm had no {\tt dbm\_close()} function, thus
there was no way to dissociate active database from a process.
A bit newer variant of dbm has the close function, and multiple
dbm's can be used.

\item[{\tt gdbm}] \mbox{}

the database is in GNU GDBM format.

\item[{\tt headers}] \mbox{}

router internal database of various headers, and how they are
to be treated.

\item[{\tt hostsfile}] \mbox{}

{\tt /etc/hosts} lookup using {\tt gethostbyname()}.

\item[{\tt incore}] \mbox{}

the database is a high-speed bundle of data kept entirely in the router process core memory.  This is for a short-term data storage, like handling duplicate detection.

\item[{\tt ldap}] \mbox{}

Mechanism for X.500 Directory access lookup with the "light-weight
directory access protocol."

\item[{\tt ndbm}] \mbox{}

the database is in NDBM (new DBM) format.

\item[{\tt ordered}] \mbox{}

the database is a text file with key-datum pairs on each line, keys are looked up using a binary search in the sorted file.

\item[{\tt selfmatch}] \mbox{}

a special type that does translate the numerical address of format
\verb/12.34.56.78/ (from within address-literal bracets) into binary
form, and checks that it is (or is not) actually our own local IP addresses.
This is used in address literal testing of addresses of type:
\verb/localpart@[12.34.56.78]/.

\item[{\tt unordered}] \mbox{}

the database is a text file with key-datum pairs on each line, keys are looked up using a sequential search.

\item[{\tt yp}] \mbox{}

Sun SunOS 4.x YP (these days "NIS") interface library.

\end{description}


A subtype is specified by appending it to the database type name separated
by a slash, or a comma.
For example, specifying {\tt bind/mx} as the argument to the `{\tt t}' option
will store `{\tt mx}' for reference by the access routines whenever a query
to that relation is processed.
The subtypes must therefore be recognized by either the database-specific
access routines (for translation into some other form), or by the database
interface itself.

For {\tt unordered} and {\tt ordered} database types, the datum 
corresponding to a particular key may be null.  This situation arises if 
the database is a simple list, with one key per line and nothing else.  
In this situation, the use of an appropriate post-processor option 
(e.g. `{\tt b}') is recommended to be able to detect whether or not 
the lookup succeeded.

The `{\tt f}' option specifies the name of the database.  This is typically a
path that either names the actual (and single) database file, or gives the
root path for a number of files comprising the database (e.g. {\tt foo} may
refer to the NDBM files `{\tt foo.pag}' and `{\tt foo.dir}').  For the 
{\tt hostsfile} type of database, the {\tt /etc/hosts} file is the one 
used (and since the normal hosts file access routines do not allow specifying 
a different file, this cannot be overridden).

The `{\tt s}' option specifies the size of the cache.  If this value 
is non-zero (by default it is 10), then an LRU cache of this size is 
maintained for previous queries to this relation, including both positive 
and negative results.

The `{\tt e}' option specifies the cache data expiration time in seconds.

The `{\tt b}' option asks that a postprocessor is applied to the database 
lookup result, so the empty string is returned from the relation query if the
database search failed, and the key itself it returned if the search
succeeded.  In the latter case, any retrieved data is discarded.  The
option letter is short for Boolean.

The `{\tt n}' option asks that a postprocessor is applied to the database 
lookup result, so the key string is returned from the relation query if the
database search failed, and the retrieved datum string is returned if the
search succeeded.  The option letter is short for Non-Null.

The `{\tt l}' option asks that all keys are converted to lowercase before 
lookup in the database.  This is mutually exclusive with the `{\tt u}' 
option.

The `{\tt u}' option asks that all keys are converted to uppercase before 
lookup in the database.  This is mutually exclusive with the `{\tt l}' 
option.

The `{\tt d}' option specifies a search routine.  Currently the only legal
argument to this option is {\tt pathalias}, specifying a driver that searches
for the key using domain name lookup rules.


\begin{figure*}
\begin{alltt}\hrule\medskip
 if [ -f /etc/resolv.conf ]; then
   relation -nt bind/cname -s 100 canon # T_CNAME canonicalize hostname
   relation -nt bind/uname uname        # T_UNAME UUCP name
   relation -bt bind/mx neighbour       # T_MX/T_WKS/T_A reachability
   relation -t  bind/mp pathalias       # T_MP pathalias lookup
 else
   relation -nt hostsfile -s 100 canon  # canonicalize hostname
   relation -t unordered -f $MAILBIN/db/hosts.uucp uname
   relation -bt hostsfile neighbour
   relation -t unordered -f /dev/null pathalias
 fi
\medskip\hrule\end{alltt}
\caption{\label{fig:adm_dbfrag1}Some examples of {\tt relation} definitions}
\end{figure*}

\begin{figure*}
\begin{alltt}\hrule\medskip
 #
 # We maintain an aliases database, and may access it via NDBM,
 # or via indirect indexing:
 #
 if [ -f $MAILBIN/db/aliases.dat ]; then
     relation -t ndbm -f $MAILBIN/db/aliases aliases
 else
     relation -it ordered -f $MAILBIN/db/aliases.idx aliases
 fi
\medskip\hrule\end{alltt}
\caption{\label{fig:adm_dbfrag2}More examples of alternate forms of
database reference}
\end{figure*}

\begin{figure*}
\begin{alltt}\hrule\medskip
 relation -t unordered -f /usr/lib/news/active -b newsgroup
 relation -t unordered -f /usr/lib/uucp/L.sys -b ldotsys
 relation -t ordered -f $MAILBIN/db/hosts.transport -d pathalias transport
\medskip\hrule\end{alltt}
\caption{\label{fig:adm_dbfrag3}More miscellaneous {\tt relation} definitions
to illustriate various possibilities}
\end{figure*}



The final argument for the {\tt relation} is not preceeded by
an option letter.
It specifies the name the relation is known under.
Note that it is quite possible for different relations to use
the same database (like in case of {\em bind}).

Some sample relation definitions are in figure \vref{fig:adm_dbfrag1}.
That fragment defines a set of relations that can be accessed in the
same way, using the same names, independent of their actual definition.

{\bf\Large CHECK! (-i option!)}
As the comment says, the relation name {\tt aliases} has
special significance to the {\em router}.
Although the relation is not special in any other way (i.e. it can
be used in the normal fashion), the semantics of the data retrieved
are bound by assumptions in the aliasing mechanism.

These assumptions are that key strings are local-name's, and
the corresponding datum gives a byte offset into another file
(the root name of the aliases file, with a {\tt .dat} extention),
which contains the actual addresses associated with that alias.

The reason for this indirection is that the number of addresses associated
with a particular alias can be very large, and this makes the traditional
simple database formats inadequate.  For example, quick lookup in a text
file is only practical if it is sorted and has a regular structure.  A
large number of addresses associated with an alias makes structuring a
problem.  The situation for DBM files and variations have problems too, due
to the intrinsic limits of the storage method.  The chosen indirection
scheme avoids such problems without loss of efficiency.

More examples on figure \vref{fig:adm_dbfrag3}, where the first
two illustrate convenient coincidences of format, and the last
definition shows what might be used if outgoing channel information
is maintained in a pathalias-format database (e.g. {\tt bar smtp!bar}
means to send mail to {\tt bar} via the SMTP channel).

\subsubsection{Using a Pathalias Database}

Accessing route databases is a rather essential capability for a mailer.
At the University of Toronto, all hosts access a centrally stored database
through a slightly modified nameserver program.
If such a setup is not practical at your site, other methods are available.
The most widespread kind of route database is produced by the {\em pathalias}
program.

{\em pathalias} generates key-value pairs of the form:
\begin{alltt}\medskip
  uunet    ai.toronto.edu!uunet!\%s
  .css.gov ai.toronto.edu!uunet!seismo!\%s
\medskip\end{alltt}
which when queried about {\tt uunet} and {\tt beno.css.gov} correspond to 
the routes:
\begin{alltt}\medskip
  ai.toronto.edu!uunet
  ai.toronto.edu!uunet!seismo!beno.css.gov
\medskip\end{alltt}

Notice that there are two basic forms of routes listed: routes to UUCP node
names and routes to subdomain gateways.  Depending on the type of route
query, the value returned from a pathalias database lookup needs to be
treated differently.
For now, this may be accomplished by a configuration
file relation definition and interface function as shown in figure
\vref{fig:adm_pathalias_driver}.

\begin{figure*}
\begin{alltt}\hrule\medskip
relation -t ndbm -f $MAILBIN/uuDB -d pathalias padb

# pathalias database lookup function
padblookup (name) \{
    local path
    path = $(padb "$name")
    tsift "$path" in
    ((.+)!)?([^!]+)!%s
        if [ "$3" == "$name" ]; then
            path = "$2!$3"
        else
            path = "$2!$3!$name"
        fi
        ;;
    .*%s.*
        echo "illegal route in pathalias db: $path"
        ;;
    tfist
    return "$path"
\}
\medskip\hrule\end{alltt}
\caption{\label{fig:adm_pathalias_driver}An example of lookup driver for
{\em pathalias} generated database}
\end{figure*}

This is actually a simplistic algorithm, but it does illustrate the method.
The lookup algorithm used when the `{\tt -d}' flag is specified in the
relation definition command is rather simple; it doesn't test various case
combinations for the keys it tries.  Therefore, the keys in the pathalias
output data should probably be converted to a single case, and the `{\tt -l}'
or `{\tt -u}' flag given in the relation definition.




\subsection{Mailing Lists and {\tt\~{}/.forward}}%
\label{mailing_list_maintenance}


Mailing lists are implemented as files in the {\tt \$MAILSHARE/lists/} 
directory (or symlinks in there to files residing elsewere, though from a system
reliability standpoint it is better to have them in that directory,
and let users have symlinks to those files -- consider the NFS
with the user home directories in other machines...)

An alternate mechanism is to implement lists in the traditional {\em sendmail}
manner, however it means feeding the message to the scheduler, and
external program ({\tt /usr/lib/sendmail}) before it comes back to the
{\em router}.

The list FILE must have protection 0664 or stricter, 0700 has invalid bits.
(ok, so the ``x"-bit is not used, but illegal it is, all the same.)
Preferrable protection is: 0600

The {\tt \$MAILSHARE/lists/} directory must be owned by root.
The directory containing the "aliases" file  ({\tt \$MAILSHARE/db/}) 
must be owned by root, and the aliases file must comply with above mentioned 
protections.

The owner of FILE gets FILE-owner, and FILE-request mails,
{\em unless any of the above listed limitations are breached}.

If FILE has protection 666 (for example), the ZMailer internal function
"{\tt \$(filepriv \$filepath)}" 
returns "{\tt \$nobody}" (userid of nobody), and  function
"{\tt \$(uid2login \$nobody)}" fails, thus losing -owner, and 
-request features.

Also lists with filepriv ``nobody'' cannot be archived simply
by having an ``address'' of form {\tt /file/path} amongst the
recipient addresses.

A mailing list is set up by creating a file in the 
{\tt \$MAILVAR/lists/} directory.
The file name is the list's name (LIST) in all lowercase
(case-insensitive matching is done by converting to lowercase
before comparison).

The file contains a list of mail addresses (typically one per line)
which are parsed to pull out the destination addresses.  This means
the users' full names can be given just as in any valid RFC822 address.

The local account which owns the file will by default receive messages
sent to LIST-owner and LIST-request.  This can be explicitly overridden
in the aliases file.  Mail to the list will go out with LIST-owner as
the sender, so list bounce messages will return to the LIST-owner
address.  Archives of the list can be created by adding a file name
address (a local pathname starting with "/") to the LIST file.  The
archive file is written with the ownership of the owner of the LIST
file.  Forwarding the mailing list into a newsgroup
can be done using a mail to news script (two generations are provided
in utils/distribute and utils/mail2news).



\subsubsection{aliases.cf Logic}

\begin{itemize}
\item
If an aliases database exists and local-part is found in it, the list of
addresses mapped to by the alias entry is substituted.
\item
If an mboxmap file exists and a mapping for the local-part is found in it,
the mapping (a ``host!homedir!user'' value) determines the remote recipient
(user@host) or recipient mailbox ({\tt homedir/../PObox/user}) if host 
is local.
\item
If local-part is a login name and a readable ``{\tt .forward}'' file 
exists in the home directory, the list of addresses it contains is
substituted.
\item
If local-part is a file basename in the {\tt \$MAILVAR/lists} 
directory, the   list of addresses contained in the file is
substituted, and the sender address set to local-part-owner.
\item
If local-part is of the form file-owner or file-request, where file is
an entry in the {\tt \$MAILVAR/lists} directory, the account name 
of the owner of the file is substituted.   (File-owner identity and correct 
protections are important.)
\item
If the local-part is of format  ``user.name'',  it is actually mapped via
separate   fullnamemap, and NEVER via {\tt \$MAILVAR/db/aliases}.
\item
If PUNTHOST is defined (in {\tt /etc/zmailer.conf}) the address
{\tt local-part@\$PUNTHOST} is substituted.  Note that in this case
the mboxmap mechanism should be used to ensure local spool
mailbox delivery for local users.
\end{itemize}


\subsubsection{aliases}

The file containing the actual aliasing data is automatically created by
the {\em router} when asked to reconstruct the aliases database.  It does this
based on a text file containing the alias definitions.  This text file,
which corresponds to the {\em sendmail} aliases file, consists of individual
alias definitions, possibly separated by blank lines or commentary.
Comments are introduced by a sharp sign (octothorp: `\#') at any point where
a token might start (for example the beginning of a line, but not in the
middle of an address), and extend to the end of the line.  Each alias
definition has the exact syntax of an RFC822 message header, containing an
address-list, except for comments.  The header field name is the local-part
being aliased to the address-list that is the header value.

The fact that an alias definition follows the syntax for an RFC822 message
header, introduces an incompatibility with {\em sendmail}.
The string ``{\tt :include:}'' at the start of a local-part
(a legacy of RFC733) has special semantics.
{\em Sendmail} would strip this prefix, and regard the rest of the local-part
as a path to a file containing a list of addresses to be included in the alias
expansion.
Indeed, the {\em router} behaves in the same manner, but because some of
the characters in the prefix are RFC822 specials, the entire local-part
must be quoted.
Thus, whereas {\em sendmail} allowed:

\begin{alltt}\medskip\hrule\medskip
people: :include:/usr/lib/mail/lists/people
\medskip\hrule\end{alltt}\medskip

the proper syntax with ZMailer is:
\begin{alltt}\medskip\hrule\medskip
people: ":include:/usr/lib/mail/lists/people"
\medskip\hrule\end{alltt}\medskip


Like {\em sendmail}, if a local-part is not found in the aliases database,
the {\em router} also checks ``{\tt \~{}local-part/.forward}''
(if such exists) for any address expansion.
The {\tt .forward} file format is also an RFC822 address list, similar
to what {\em sendmail} expects.

As special cases, a local-part starting with a pipe character (`\verb/|/')
is treated as mail destined for a program (the rest of the local-part is any
valid argument to a ``{\tt sh -c}'' command), and a local-part starting with
a slash character (`{\tt /}') is treated as mail destined for the file named
by the local-part.

General format:

\begin{alltt}\medskip\hrule\medskip
 local-address-token:    "replacement address" ,
                         "extension line w/ another addr"
 --------------------    --------------------------------
     "The Key"                "The Data"
\medskip\hrule\end{alltt}\medskip

Protection of the aliases database must be at least 0644.
Protection of the {\tt \$MAILVAR/db/} directory must be at least 03755.



The following processing is done for (replacement) local-parts
(local mail addresses):  Note that @'s are not allowed in any local-part.



If the local-part starts with ``\verb/|/'' assume it is a command
specification:
\begin{alltt}\medskip\hrule\medskip
prog-pipe: "|/path/to/program -args"
\medskip\hrule\end{alltt}\medskip

If the local-part starts with ``/'' assume it is a file pathname:
\begin{alltt}\medskip\hrule\medskip
file-path: "/path/to/file"
\medskip\hrule\end{alltt}\medskip

If the local-part starts with ``{\tt :include:}'' the rest should be a file
pathname of a list of mail addresses.  They are substituted:
\begin{alltt}\medskip\hrule\medskip
included-list: ":include:/path/to/address/file"
\medskip\hrule\end{alltt}\medskip


After this point, all matches are case-insensitive by means of translating
the value to be looked up to lower-case, and then conducting a case-sensitive
lookup. {\em All keys in aliases et.al. must be in lower case -- you
can achieve this with bundled ``{\tt newaliases}'' script, which calls
``{\tt makedb}'' with ``-l'' option to lowercasify keys.}.
(The hash functions inside  {\tt ndbm/gdbm/db/dbm}  are case sensitive, and 
as such, there is no way to avoid this requirement.)


\subsubsection{About Large Lists, and Memory Consumption}

With old configuration scripts there used to be problems with list
expansion causing serious memory bloat (of $O(n^{2})$ kind of memory
consumption\ldots)
{\em ZMailer-2.98} did introduce a working solution via the builtin
{\tt\$(listexpand ..)} function.

The rest of these notes apply to older config files using old style pipe
of:
\begin{alltt}\medskip\hrule\medskip
 $(listaddresses < file | maprrouter ...)
\medskip\hrule\end{alltt}\medskip
where the maprrouter caused the actual memory bloat...

The primary reason behind this is that the router does nnecessary
copies of objects -- it uses memory in a form of heap storage which is
allocated progressively and managed via ``{\tt mark()}'', and
``{\tt release()}'' type calls.
The ``{\tt release()}'' calls are done when exiting ``{\em zmsh}''
interpreter, and never during script running (however calling
builtin C functions which call scripts is another story -- there
are ``{\tt release()}'' checkpoints available then.)

Especially bad thing for memory use to do is to use ``{\tt setf}'' function.

Internal list expansion (through the  {\tt \$MAILVAR/lists/LISTNAME} 
mechanism) is a sure way to expand {\em router} process memory usage.

You can decrease the memory requirement dramatically, if you can
feed all the addresses in the envelope, or via {\tt utils/listexpand.c}
utility (alpha-test tool on 1-Sep-94) You don't need to worry about 
it unless your list is 100+ recipients, only then the memory usage starts 
to bloat seriously with the old-style in-core {\tt \$(listaddress ...)} 
expander.

Although more interesting and useful models exist, the mail forwarding
functionality of ZMailer has been designed to generally emulate the
interface and behaviour of {\em sendmail}.
The mechanisms that accomplish this are likely to be generalized in
a future version.


\subsubsection{Security Considerations}



A LIST file must not be world writable, while most likely it can be
group-writable.
The {\tt \$MAILVAR/lists/} directory must also not be group or world
writable and must be owned by root or by the owner of the LIST file.
Otherwise the file is declared insecure and all addresses in the file
get the least possible privilege associated (the ``nobody'' uid).
This can cause various things to break, for example mailing list archival,
or the -owner and -request features if ``nobody'' is not a valid account.

There is a mechanism to override using the modes on a {\tt file/directory}
as an indicator of its safeness.

Turning on the sticky bit on a file or directory tells the mailer to treat
it as if it was only owner writable independent of its actual modes.

This allows {\tt MAILVAR/lists/} to be group or world writable and
sticky-bitted if you want your general user population (or special admin
group) to be able to create mailing lists.

%\end{multicols}